### 2017-12-11 Writing/disparity_review.tex
Added methods to review

Biological data are complex; understanding the ecology and evolution of species often requires that we analyse multiple variables that covary with each other, and through space and time.
One solution to this problem is to analyse data in a multivariate framework.
Multivariate analyses aim to capture and incorporate this multidimensional complexity, while providing outputs that are interpretable in a physical world of only three dimensions.
Classical multidimensional species features that have been analysed through multivariate data ranges from their morphology \citep{raup1966geometric}, to their functional traits \citep{diaz2016global}.
However, such analysis are not limited at the species level and can also be applied to ecosystems \citep{DonohueDim}.

Key multidimensional features of species that have important roles in ecology and evolution include morphology, functional traits. etc. etc.
Many of these can be represented as matrices, these can be used in our package.
In the interests of clarity/brievety, we here focus on just one kind of data: morphological diversity.

Furthermore, we will also make a clear distinction between the multidimensional space which is a specific mathematical object and the disparity, which is a metric describing or summarising one or more aspects of this space.

Both can be defined in various ways depending on the authors, for example, disparity is defined as the weighted mean pairwise dissimilarity in \cite{Close2015} or the ellipsoid volume in \cite{DonohueDim} and the multidimensional space is defined as the morphospace in \cite{raup1966geometric} and the morpho-functional space in \cite{diaz2016global}.

The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.

This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.

Disparity is an old concept blabalbal

\section{What \textit{is} disparity}

From a semantic view: morphospace or aspect of the morphospace?

From a biological view: what does an increase or a decrease in disparity represent?

\section{Why do we need disparity}

@@

@@

Methods?
Data?

@@

Yay!

% \section{Data availability and reproducibility}
% Data are available on Dryad or Figshare.
% Code for reproducing the analyses is available on GitHub (\url{github.com/TGuillerme/SpatioTemporal_Disparity}).

\section{Acknowledgments}
Royal Society..
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

@@

>>>>

%§1: disparity is observed everytwhere
Disparity is an old concept blabalbal coming from the observation that some groups of species are more similar/dissimilar than other.

It can be also applied between groups as a concept of ecological niches: a group occupying many niches will have a stronger disparity than a groups with less niches.

It is decoupled from diversity.

\bigskip
%§2: "phylogeny" of disparity

In palaeobiology, morphological disparity (the (dis)similarities between the morphologies of groups of organisms) concept as a tool for macroevolutionary studies stems to several seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.

This concept was then used in geometric morphometric and expanded to macroevolutionary studies in general (Harmon's dtt).

\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.

\bigskip

%§3: a very popular concept

Disparity is really popular (Fig. \ref{Fig:GoogleOccurences}):

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/GoogleScholarOccurences.pdf} 
\caption{Number of papers on Google Scholar matching the search ``morphological disparity'' per year. In black, the match is in the paper and in grey, in the title. We collected the number of matches per year from 1960 to 2016 in Google Scholar for the terms "morphological disparity" both in the text (fuzzy matching) or in the title (exact matching). The data was collected on the 1st of November 2017.}
\label{Fig:GoogleOccurences}
\end{figure}

\section{Why do we need disparity}
Which macroevolutionary questions are tackled through disparity analysis?

@@

Disparity can be 
 -From a semantic view: morphospace or aspect of the morphospace?
 -From a biological view: what does an increase or a decrease in disparity represent?

But in the end, what \textit{is} disparity

\section{But what \textit{is} disparity?}

Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}

% §Methods for data:
\subsection{Data collection methodology}
To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
For the 230 relevant papers among the 500 matches, we collected the following methodological data:

\begin{itemize}
    \item What was the focal biological group?
    \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
    \item Was data collected on the full organism or not?
    \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
    \item How was the disparity metric(s) explicitly defined?
    \item Which statistical test was applied to test the disparity related hypothesis?
    \item Was phylogeny taken into account or not?
\end{itemize}

We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g. a disparity metric was measured but not described anywhere in the paper).

The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).

To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in the following categories:

[DESCRIBE THE CATEGORIES HERE (from \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd})]

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
\caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
\label{Fig:MethodsProportions}
\end{figure}

[ADD Biological group, full organism and phylogeny in the supplementary]

\subsection{Disparity data}
Describing the many data inputs (from collected data to morphospace):

\subsection{Disparity metrics}
Describing the many metrics (what is actually measured):

\subsection{Disparity hypothesis}
Describing the many outputs (what and how is it tested):

\subsection{The three main disparity analysis}
From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.

\begin{itemize}
    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

    \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA). Disparity is often measured as a distance metric (e.g. the distance between the species and a point in the morphospace such as its centroid). Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).

    \item \textbf{The ``\texttt{dtt}'' approach:} continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.

\end{itemize}

Of course some studies use a combination of these three methods or none of them at all!

Also, among each category XX\% of studies use multiple approaches.

[ADD Biological group, full organism and phylogeny in the supplementary]

@@

How to compare disparity between groups?

How to compare disparity between methods?

Hot to compare disparity between disciplines?

%TG: we can find solutions from other fields
It bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.

@@

A quick guideline for good disparity analysis:

Maybe we need something like in Parham et al 2012 (Best Practices for Justifying Fossil Calibrations): an easy an identifiable description of the pipeline containing: 1) the type of data, 2) the morphospace and 3) the metric?

\section{Acknowledgments}
The Royal Society.
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

@@

% The multidimensional space can be defined in many ways and arise from many mathematical transformations of the data such as the pairwise distance matrix \citep{Close2015}, a principal coordinates analysis \citep[PCO;][]{Brusatte12092008}, a principal components analysis \citep[PCA;][]{zelditch2012geometric}, a multidimensional scaling \citep[MDS;][]{DonohueDim}, etc.
% Similarly, disparity metrics \citep[or indices;][]{Hopkins2017} can defined in many ways \citep[e.g.][or combinations thereof]{Wills2001,Ciampaglio2001,foth2012different,DonohueDim,Hughes20082013,finlay2015morphological,Close2015,diaz2016global}.
% Finally, difference between disparity metrics can also be measured in many ways: using NPMANOVA \citep[e.g.][]{Brusatte12092008}, multidimensional permutation test \citep[e.g.][]{diaz2016global} or even simple confidence interval overlap \citep[e.g.][]{halliday2016eutherian}.

% This variety of definitions and analysis have been developed in an equal variety of softwares such as \texttt{GINGKO} in javascript \citep{bouxin2005ginkgo,de2007ginkgo} or \texttt{geomorph} \citep{adams2013geomorph,adams2017geometric}, \texttt{Claddis} \citep{Claddis}, or \texttt{vegan} \citep{oksanen2007vegan} in R \citep{R}.
% This results in the need to learn different languages (or at least - when restricted to R - different packages with different standards) as well as making analysis sometimes idiosyncratic and often complex to repeat since they are based on a particular feature from a particular software.
% For example, in the excellent and widely used \texttt{geomorph} package morphological disparity analysis can be ran using the \texttt{morphol.disparity} function.
% Unfortunately, however, the multidimensional space can only be defined as the ordination of the procrustes transform of geometric morphometric landmarks, the disparity can only be defined as the Procrustes variance and the difference between groups can only be measured through permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.

% In all these analysis, each set of multivariate traits forms a multidimensional space.
% \hl{This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).}
% \hl{These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (}
% \citealt{lloyd2016estimating}
% \hl{and references therein; e.g. in}
% \citealt{Close2015}
% \hl{), or as outputs from an ordination, whether it being a principal components analysis (PCA,}
% \citealt{PCA}
% \hl{; e.g. in}
% \citealt{zelditch2012geometric}
% \hl{), a metric scaling (PCO, PCoA,}
% \citealt{PCO}
% \hl{; e.g. in}
% \citealt{Brusatte12092008}
% \hl{) or a non-metric scaling (MDS, NMDS,}
% \citealt{MDS}
% \hl{; e.g. in}
% \citealt{Liow2004,DonohueDim}
% \hl{).}
% The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
% For example, when using morphological traits the multivariate space will be a morphospace, when using ecological traits it may be referred to as an ecospace or trait space.

% \hl{It is then possible to measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
% This requires the definition of a proxy for space occupancy: the disparity metric}
% \citep[or index;][]{Hopkins2017}
% \hl{which can be measured in a multitude of ways.}
% \hl{For example, one could use a metric based on the variance or the range of each axis of space}
% \citep{Wills2001, Ciampaglio2001}
% \hl{, a distance (e.g. Euclidean) measured between observations}
% \citep{foote1993contributions,Foote29111996}
% \hl{, a more direct approximation of the hyper volume }
% \citep{cornwell2006trait,DonohueDim}
% \hl{, or many more }
% \citep[e.g.][]{navarro2003MDA}.

% Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
% \hl{NPMANOVA, }
% \citealt{NPMANOVA}
% \hl{; e.g. in }
% \citealt{Brusatte12092008}
% \hl{)}
% multidimensional permutation tests
% \hl{(}
% \citealt{ManlyPermutations}
% \hl{; e.g. in }
% \citealt{diaz2016global}
% \hl{)}
% or even simply by looking at the confidence interval overlaps between disparity measurements
% \hl{().}
% In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.
